---
title: "Lab_R_SQL_dplyr"
author: "Gabriele Etta"
format: html
editor: visual
---

# Lab 10/4/2024

In today's lab we will cover the following topics

-   Setup a *duckDB* database
-   Interact with a *duckDB* database using R
-   Manage dataframes using *dplyr* package
-   Visualize results with *ggplot2*

To replicate this lab, duckDB needs to be installed on your laptop.

# Setup libraries and connections

We begin by installing the required packages

```{r}
install.packages("duckdb")
install.packages("dplyr")
install.packages("ggplot2")
library(dplyr, quietly = T)
library(duckdb, quietly = T)
library(ggplot2, quietly = T)
```

Next, we connect to the duckDB database previously created

```{r, connect to duckDB}
# Create a DUCKDB database called titanic
con <- dbConnect(duckdb::duckdb(), dbname = "titanic.db")

# Load titanic data
titanic_data <- read.csv("titanic.csv")

# Create a titanic data and add the tuples from the CSV file 
dbWriteTable(con, "passengers", titanic_data, row.names = FALSE)
```

Get the adult passengers from *passengers* table

```{r, performing SQL query}
df_adult_passengers <- dbGetQuery(con, "SELECT * FROM passengers WHERE age > 18")

df_adult_passengers <- as.data.frame(df_adult_passengers)
df_adult_passengers
```

We now filter the adult passengers by using dplyr, keeping only those adults who survived

```{r, filtering with dplyr}
  filtered_data <- df_adult_passengers %>%
                    filter(Survived == 1 & Age > 18)
  
  # Compute average
  mean_age <- filtered_data %>%
                summarize(mean_age = mean(Age, na.rm = TRUE))
  
  mean_age
```

Finally, we visualize distribution and survival rates using ggplot2

```{r, visualize plots}

age_plot <- ggplot(filtered_data, aes(x = Age)) +
              geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
              labs(title = "Age Distribution of Surviving Passengers over 18",
                   x = "Age",
                   y = "Frequency")

# Plot survival rate by passenger class
class_survival_plot <- ggplot(titanic_data, aes(x = factor(Pclass), fill = factor(Survived))) +
                          geom_bar(position = "dodge", color = "black") +
                          scale_fill_manual(values = c("red", "green")) +
                          labs(title = "Survival Rate by Passenger Class",
                               x = "Passenger Class",
                               y = "Count",
                               fill = "Survived")

# Display plots
print(age_plot)
print(class_survival_plot)
```
